---
title: "Athlete Training Analysis: Fast-Twitch vs. Slow-Twitch"
author: "Alex"
date: "May 2, 2025"
output: html_document
---

## Introduction
This project compares athletes training fast-twitch muscle fibers (sprinting, resistance training, plyometrics) versus those training slow-twitch fibers (jogging, light running). We analyze speed, endurance, and recovery time using a simulated dataset inspired by real-world studies.

## Data Overview
The dataset includes 100 athletes with the following variables:

- **Training_Type**: Fast-Twitch or Slow-Twitch
- **Age**: 18–40 years
- **STMF_Percentage**: Slow-twitch muscle fiber percentage
- **Speed_mps**: Speed in meters per second (100m sprint)
- **Endurance_min**: Endurance in minutes (sustained jogging)
- **Recovery_Time_min**: Recovery time after high-intensity effort
- **Training_Frequency**: Weekly training sessions

```{r setup, include=FALSE}
library(tidyverse)
```

```{r load-data, echo=TRUE}
data <- read_csv("athlete_training_data_cleaned.csv")
```
## Data Analysis

### Summary Statistics

```{r summary-stats, echo=TRUE}

summary_stats <- data %>%
  group_by(Training_Type) %>%
  summarise(
    Avg_Speed = mean(Speed_mps),
    Avg_Endurance = mean(Endurance_min),
    Avg_Recovery = mean(Recovery_Time_min),
    Avg_STMF = mean(STMF_Percentage)
  )
summary_stats
```

### Statistical Tests

```{r t-tests, echo=TRUE}

speed_test <- t.test(Speed_mps ~ Training_Type, data = data)
endurance_test <- t.test(Endurance_min ~ Training_Type, data = data)
recovery_test <- t.test(Recovery_Time_min ~ Training_Type, data = data)
speed_test
endurance_test
recovery_test
```

## Data Visualizations

### Speed Comparison

```{r speed-plot, echo=TRUE, fig.align='center'}

speed_plot <- ggplot(data, aes(x = Training_Type, y = Speed_mps, fill = Training_Type)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Speed Comparison by Training Type", x = "Training Type", y = "Speed (m/s)") +
  scale_fill_manual(values = c("Fast-Twitch" = "coral", "Slow-Twitch" = "skyblue"))
speed_plot
```

### Endurance Comparison

```{r endurance-plot, echo=TRUE, fig.align='center'}

endurance_plot <- ggplot(data, aes(x = Training_Type, y = Endurance_min, fill = Training_Type)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Endurance Comparison by Training Type", x = "Training Type", y = "Endurance (min)") +
  scale_fill_manual(values = c("Fast-Twitch" = "coral", "Slow-Twitch" = "skyblue"))
endurance_plot
```

### Speed vs. Endurance

```{r scatter-plot, echo=TRUE, fig.align='center'}

scatter_plot <- ggplot(data, aes(x = Speed_mps, y = Endurance_min, color = Training_Type)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Speed vs. Endurance by Training Type", x = "Speed (m/s)", y = "Endurance (min)") +
  scale_color_manual(values = c("Fast-Twitch" = "coral", "Slow-Twitch" = "skyblue"))
scatter_plot
```

## Tableau Dashboard
Explore an interactive dashboard with additional visualizations: https://public.tableau.com/app/profile/alexander.gallagher/viz/Athlete_Training_Comparison/Dashboard1

## Key Findings
Fast-twitch athletes have higher speeds (8–10 m/s) compared to slow-twitch athletes (5–6.7 m/s).  

Slow-twitch athletes excel in endurance (60–90 minutes) compared to fast-twitch athletes (30–50 minutes).  

Fast-twitch athletes take longer to recover (40–60 minutes) than slow-twitch athletes (20–40 minutes).

## Conclusion
Training focus significantly impacts athlete performance. Fast-twitch training enhances speed, while slow-twitch training boosts endurance. These insights can guide training programs for athletes based on their goals.



